---
title: "Impervious Surfaces - Step 3 Coastal Impact Raster"
output: html_document
date: "2026-02-22"
---
# Summary

NOTE: THIS WAS COPIED TO THE NEW STANDALONE URBAN REPO BY MADDIE 08-10-2026. THIS VERSION WILL NO LONGER BE UPDATED

This markdown walks through the  model used to quantify the impervious surfaces have on nearshorewaters. 

The first part of creating this driver has been moved to Python, due to the large size of the input data (3 and 1 m rasters). The workflow is as follows:
- Py script 1: Mosaic 2011 and 2021 together, and mask 2011 by 2021 to create two new versions. 
- Py script 2: Summarize the number of impervious pixels per watershed


Inputs to this workflow, Step 3, are as follows

- Aggregated pixel count tables, which have number of pixels for each watershed for each TILE of each year (9 tiles per year). There is one table for the counts from the mosaicked 2021 version and one from the masked 2011 version
- Vector file delineating watersheds

## Working notes:
- Need to fix file paths here to point to outputs folder for tables from step 2
- Need to fix the set up document to point to new repository once its created


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

#create new directory for this folder (this needs to be changed)
urban_dir <- here("2_Urban_Runoff")

# set up packages, file paths and other common workflow items
source(here("workflow/set_up.R"))

# create directory for urban folder on dropbox
urban_dropbox <- file.path(driver_dir,"2_Urban_Runoff")
```



## Part 1: Summarize Area of Impervious Surface per Watershed

### Previous versions 1.2 - 1.5

```{r}
# function to process each csv version we make

process_urban_pixels <- function(file_path, year_label) {
  read.csv(file_path) %>%
    # Pivot and summarize
    pivot_longer(cols = contains("pixel"), names_to = "raster", values_to = "pixel_count") %>%
    group_by(zone_id) %>%
    summarize(
      pixels = sum(pixel_count, na.rm = TRUE),
      area_km2 = (pixels * 5.76) / 1e6, # Consistent calc
      .groups = "drop"
    ) %>%
    # Rename columns to include the year (e.g., pix_2011, area_2011)
    rename_with(~paste0(.x, "_", year_label), c(pixels, area_km2))
}

# apply to csvs

# --- Version 1.5 ----

v15_21 <- process_urban_pixels(file.path(urban_dir, "outputs/aggregated_pixel_counts_2021_newWurbanmask_v1.5.csv"), "21")
v15_a <- process_urban_pixels(file.path(urban_dir, "outputs/aggregated_pixel_counts_2011_newWurbanmask_v1.5.csv"), "11")
v15_mauinui <- process_urban_pixels(file.path(urban_dir, "outputs/aggregated_pixel_counts_2011_newWurbanmask_v1.5_mauiNui.csv"), "11")
v15_11 <- full_join(v15_a,v15_mauinui, by = c("zone_id")) %>% 
  mutate(pixels_11.x = ifelse(pixels_11.x == 0, pixels_11.y,pixels_11.x)) %>% 
  mutate(area_km2_11.x = ifelse(area_km2_11.x == 0.00000000, area_km2_11.y,area_km2_11.x)) %>% 
  select(zone_id, pixels_11 = pixels_11.x, area_km2_11 = area_km2_11.x)


# Load mapping and watershed data
zone_id_mapping <- read.csv(file.path(urban_dir, "zone_id_mapping.csv")) %>% rename(zone_id = zone_int) %>% mutate(HUC12 = as.character(HUC12))
watershed_meta <- read_sf(file.path(common_dir, "Common_2024/Finalized/Coastlines_Watersheds/WBD_HU12_n83utm4_2023fixed/WBD_HU12_n83utm4_2023fixed.shp")) %>%
  select(HUC12, Name, AreaSqKm) %>% 
  mutate(HUC12 = as.character(HUC12))

islands <- read_sf(file.path(dropbox_dir, "Drivers/1_AgGolf_Runoff/outputs/watershed_level/polys/watershed_allAg_clean.shp")) %>%
  select(HUC12, island) %>% 
  st_drop_geometry() 

# Combine into one lookup table
lookup_table <- watershed_meta %>%
  left_join(zone_id_mapping, by = "HUC12") %>%
  left_join(islands, by = "HUC12")
  

# Join years together 
table_v1.5 <- lookup_table %>%
  left_join(v15_11, by = "zone_id") %>%
  left_join(v15_21, by = "zone_id")

table_v1.5_diff <- table_v1.5 %>% 
  mutate(diff11_21 = area_km2_21 - area_km2_11)

hist(table_v1.5_diff$diff11_21)

# save difference calculation

st_write(table_v1.5, file.path(dropbox_dir,"/Drivers/2_Urban_Runoff/output_data/watershed_level/watersheds_wImpervious_2021_2011_updated0326_v1.5.shp"))

```


### Visualize Difference Between Years

```{r viz-years-diff }


# visualize with maps
library(tidyverse)
library(ggplot2)
library(viridis)
library(purrr)
library(patchwork)


mapping_df <- table_v1.5_diff %>% 
  pivot_longer(cols = starts_with("diff"),
               names_to = "time_difference",
               values_to = "change_value") %>% 
  mutate(fill_value = ifelse(change_value > 0, 1, change_value))


# map code

oahu_map <- mapping_df %>% filter(island == "Oahu")
ggplot(oahu_map)+
  geom_sf(aes(fill = change_value), color = "grey", size = 0.1)+
  scale_fill_gradient2(low = "coral", 
                       mid = "white",
                       high = "darkgreen",
                       midpoint = 0)+
  labs(title = "Change in Impervious Surface per Watershed Over Time", 
       fill = "Change (km2)")+
  theme_minimal()


# loop through each island

islands <- unique(mapping_df$island)

plot_list <- map(islands, function(isl){
  
  island_data <- mapping_df %>% filter(island == isl)
  
  ggplot(island_data)+
    geom_sf(aes(fill = change_value), color = "grey", size = 0.1)+
    scale_fill_gradient2(low = "coral",
                         mid = "white",
                         high = "darkgreen",
                         midpoint = 0)+
  labs(fill = "Change (km2)")+
  theme_minimal()
  })

# combine into a grid
combined_map <- wrap_plots(plot_list) + 
  plot_annotation(title = "Watershed Changes Across All Islands")
combined_map
ggsave("hawaii_watersheds_grid_v1.5.png", combined_map, width = 12, height = 8, dpi = 300)


# if you want to do separate maps?

# create maps

ggplot(maui_heatmap)+
  geom_sf(aes(fill = change_value), color = "grey", size = 0.1)+
  scale_fill_gradient(low = "white", high = "darkgreen")+
  labs(title = "Change in Impervious Surface per Watershed Over Time (Maui)", 
       fill = "Change (km2)")+
  theme_minimal()

ggsave("maui_urban_change_new1.5.png", 
       bg = "white")

ggplot(oahu_heatmap)+
  geom_sf(aes(fill = change_value), color = "grey", size = 0.1)+
  scale_fill_gradient(low = "white", high = "darkgreen")+
  labs(title = "Change in Impervious Surface per Watershed Over Time (Oahu)", 
       fill = "Change (km2)")+
  theme_minimal()

ggsave("oahu_urban_change_new1.5.png", 
       bg = "white")

ggplot(kauai_heatmap)+
  geom_sf(aes(fill = change_value), color = "grey", size = 0.1)+
  scale_fill_gradient(low = "white", high = "darkgreen")+
  labs(title = "Change in Impervious Surface per Watershed Over Time (Kauai)", 
       fill = "Change (km2)")+
  theme_minimal()

ggsave("kauai_urban_change_new1.5.png", 
       bg = "white")

ggplot(hawaii_heatmap)+
  geom_sf(aes(fill = change_value), color = "grey", size = 0.1)+
  scale_fill_gradient(low = "white", high = "darkgreen")+
  labs(title = "Change in Impervious Surface per Watershed Over Time (HI)", 
       fill = "Change (km2)")+
  theme_minimal()

ggsave("hawaii_urban_change_newv1.5.png", 
       bg = "white")


```
After re-running 2011 with xarray method, the numbers make a bit more sense. Still worth "spot checking" the watersheds with more impervious surface in 2011 than 2021. But not as many.
I also ended up resampling the 2017 and 2021 to match. Sadly we still see the same issue for a lot of watersheds, but its getting better. 

### Last check: Normalized values

While the trends we see above aren't great, what matters for the model is the change in watersheds relative to each other. Therefore we can normalize by the largest value and see if the location of the most impacted watersheds has shifted at all. I'll do this by island to start and then at the state level.


```{r normalized-viz}

# Get max values for each year

max_df <- data.frame(
  max_2011 = max(years_compare$areaImp_11_km),
  max_2021 = max(years_compare$areaImp_21_km)
)

all_years_max <- max(max_df) # the maximum of all, which as of July 2025 is still 2011

# normalize all by the biggest
years_compare_norm <- years_compare %>% 
  select(HUC12, Name, island, areaImp_11_km, areaImp_21_km) %>% 
  mutate(across(starts_with("area"), ~.x/all_years_max, .names = "{.col}_norm"))


# Visualize

heatmap_df_norm <- years_compare_norm %>% 
  pivot_longer(cols = ends_with("norm"),
               names_to = "time_step",
               values_to = "normalized_value") %>% 
  #filter(change_value < 0.1 | change_value > -0.1) %>%  # only 24 rows were zero, but a lot were less than a quarter km2
  mutate(time_step = recode(time_step,
                                 areaImp_11_km_norm = "2011",
                                 areaImp_21_km_norm = "2021"
                                 ))
  #mutate(fill_value = ifelse(change_value > 0, 1, change_value))



maui_norm <- heatmap_df_norm %>% filter(island == "Maui")
oahu_norm <- heatmap_df_norm %>% filter(island == "Oahu")
kauai_norm <- heatmap_df_norm %>% filter(island == "Kauai")
hawaii_norm <- heatmap_df_norm %>%  filter(island == "Hawaii")



ggplot(maui_norm)+
  geom_sf(aes(fill = normalized_value), color = "grey", size = 0.1)+
  scale_fill_gradient(
    low = "white",
    high = "steel blue",
    limits = c(min(maui_norm$normalized_value, na.rm = TRUE),
               max(maui_norm$normalized_value, na.rm = TRUE)),
    name = "Normalized Area"
  )+
  #scale_fill_gradient2(low = "coral", mid = "white", high = "seagreen", midpoint = 0) + 
  facet_wrap(~time_step) +
  labs(title = "Normalized Impervious Surface Maui", 
       fill = "Area (km2)")+
  theme_minimal()


ggplot(oahu_norm)+
  geom_sf(aes(fill = normalized_value), color = "grey", size = 0.1)+
  scale_fill_gradient(
    low = "white",
    high = "steel blue",
    limits = c(min(oahu_norm$normalized_value, na.rm = TRUE),
               max(oahu_norm$normalized_value, na.rm = TRUE)),
    name = "Normalized Area"
  )+
  #scale_fill_gradient2(low = "coral", mid = "white", high = "seagreen", midpoint = 0) + 
  facet_wrap(~time_step) +
  labs(title = "Normalized Impervious Surface Oahu", 
       fill = "Area (km2)")+
  theme_minimal()



ggplot(kauai_norm)+
  geom_sf(aes(fill = normalized_value), color = "grey", size = 0.1)+
  scale_fill_gradient(
    low = "white",
    high = "steel blue",
    limits = c(min(kauai_norm$normalized_value, na.rm = TRUE),
               max(kauai_norm$normalized_value, na.rm = TRUE)),
    name = "Normalized Area"
  )+
  #scale_fill_gradient2(low = "coral", mid = "white", high = "seagreen", midpoint = 0) + 
  facet_wrap(~time_step) +
  labs(title = "Normalized Impervious Surface Kauai", 
       fill = "Area (km2)")+
  theme_minimal()


ggplot(hawaii_norm)+
  geom_sf(aes(fill = normalized_value), color = "grey", size = 0.1)+
  scale_fill_gradient(
    low = "white",
    high = "steel blue",
    limits = c(min(hawaii_norm$normalized_value, na.rm = TRUE),
               max(hawaii_norm$normalized_value, na.rm = TRUE)),
    name = "Normalized Area"
  )+
  #scale_fill_gradient2(low = "coral", mid = "white", high = "seagreen", midpoint = 0) + 
  facet_wrap(~time_step) +
  labs(title = "Normalized Impervious Surface Hawaii Island", 
       fill = "Area (km2)")+
  theme_minimal()
```

## Part 2: Disperse values to nearshore

### Goal

The second part of this model uses spatial operations to extend the impact metric, in this case, area of imperivous surface, into the nearshore environment.

### Assumptions
- Runoff only likely affects ocean space that is **within 2 km of the shoreline**
- Watersheds impact nearshore area that is closes to them. Said another way, nearshore areas are most affected by whichever watershed is closest, in distance, to them

Note: the first function I've created for these steps rely on both `terra` and `raster`, because I use the `fasterize::fasterize` tool which is really helpful. As of [this](https://predictiveecology.org/posts/2024-07-03-terra%20transition.html) blog post, there does not seem to be an equivalent tool for `terra`. Something to explore in case this starts breaking. 

Inputs: 

- Watershed shapefile with attribute for area of impervious surface - from output_data/watershed_level
- Ocean Mask (LBSP, both years)

Outputs: Dropbox scarch folder: scratch_data/focStat_urban_",year,.tif

```{r}

# read in shapefile with impervious surface

watersheds <- read_sf(file.path(dropbox_dir,"/Drivers/2_Urban_Runoff/output_data/watershed_level/watersheds_wImpervious_2021_2011_updated0326_v1.5.shp"))

# Load custom "rasterize_expand" function 

source(file.path(urban_dir,"R/Part2_function.R")) # function

# if re-running 2011, use pre 2018!
ocean_mask <- raster(file.path(dropbox_dir,"Drivers/0_Input_Data_and_Models/Common_2024/Finalized/Masks/OceanMask_LBSP_post2018/OceanMask_LBSP_post2018.tif"))
#ocean_mask <- raster(file.path(dropbox_dir,"Drivers/0_Input_Data_and_Models/Common_2024/Finalized/Masks/OceanMask_LBSP_pre2018/OceanMask_LBSP_pre2018.tif"))

ocean_mask_spat <- rast(ocean_mask)
par(mar = c(3, 3, 2, 2))
plot(ocean_mask_spat) # check it out

e <- ext(ocean_mask)
origin <- origin(ocean_mask)

ocean_mask_nad83utm <- project(ocean_mask_spat,"EPSG:26904")
crs_ocean_mask <- crs(ocean_mask_nad83utm)

## project watershed file from part 1

watershed_nad83utm <- st_transform(watersheds, crs = crs_ocean_mask)

# apply to 2011 and 2021

focalStat_2011 <- rasterize_expand("ar_2_11","2011")
focalStat_2021 <- rasterize_expand("ar_2_21","2021")

plot(focalStat_2011)
plot(focalStat_2021)

```


## Part 3: Apply Decay Function

### Goal 
This part of the model, applies a decay function to model a diminishing impact of runoff as distance from shore increases. 

### Assumptions
Impact decreases with distance from shore.

### Equation: 

> **runoff** * Exp(-1 * (**Distance_From_Shore** ^ 2.5) / 10 ^ 7.5)

Inputs: Outputs from Step2, in scratch_data
Output: Final urban runoff coastal impact for each raster, in "output_data" 

```{r}

# Load custom offshore decay function "offshore_decay"

source(file.path(urban_dir,"R/Part3_function.R")) #function


# read in Euclidean distance raster. It is also in Nad83 Utm4 

dist_from_shore <- terra::rast(file.path(dropbox_dir,"/Drivers/0_Input_Data_and_Models/Common_2024/Finalized/EucDistance/DistFromSHore_Euc_2023/DistFromSHore_Euc_2023.tif"))

# read in raster from step 2, if you are starting here. Note: this kind of messes up the numbers maybe?

focalStat_2011 <- terra::rast(file.path(urban_dropbox,"scratch_data/focSt_extract_urban_2011.tif"))
focalStat_2021 <- terra::rast(file.path(urban_dropbox,"scratch_data/focSt_extract_urban_2021.tif"))

# apply decay function 

final_urban2011 <- offshore_decay(focalStat_2011, dist_from_shore,"2011")
final_urban2021 <- offshore_decay(focalStat_2021, dist_from_shore,"2021")


```


### Visualize Results & Save

```{r quick-plots }

#par(mfrow = c(1,3))
plot(final_urban2011)

plot(final_urban2021)

minmax(final_urban2021)

minmax(final_urban2011)

```

```{r map}
library(ggnewscale)
# read in raster to map

urban_v1.2 <- rast(file.path(urban_dropbox,"output_data/finalized_Urban/V1.5/UrbanRunoff_CCAP_2021_01.tif"))
urban_v1.2_f <- terra::as.data.frame(urban_v1.2, xy = TRUE,na.rm = TRUE)
colnames(urban_v1.2_f)[3] <- "urban"

# read in basemaps
source(here("workflow/3D_basemaps.R"))

# buildmap

oahu_with_dem + 
  new_scale_fill()+ # reset scale
  geom_raster(data = urban_v1.2_f, aes(x = x, y = y, fill = urban), alpha = 0.7)+
  scale_fill_gradientn(colors= c("aquamarine3","orange"))


```


## QAQC 1: Check resolution, extent and alignment

```{r}

# Read in reach raster you want to QAQC

final_2021 <- terra::rast(file.path(urban_dropbox,"/output_data/finalized_Urban/V1.5/UrbanRunoff_CCAP_2021_decay_masked.tif"))
  
  
final_2011 <- terra::rast(file.path(urban_dropbox,"/output_data/finalized_Urban/V1.5/UrbanRunoff_CCAP_2011_decay_masked.tif"))


# Read in ocean mask 

ocean_mask_post2018 <- terra::rast(file.path(dropbox_dir,"Drivers/0_Input_Data_and_Models/Common_2024/Finalized/Masks/OceanMask_LBSP_post2018/OceanMask_LBSP_post2018.tif"))

ocean_mask_pre2018 <- terra::rast(file.path(dropbox_dir,"Drivers/0_Input_Data_and_Models/Common_2024/Finalized/Masks/OceanMask_LBSP_pre2018/OceanMask_LBSP_pre2018.tif"))

# Compare outputs to eachother
terra::compareGeom(final_2021, final_2011)
# Compare each output to the ocean mask
terra::compareGeom(final_2021, ocean_mask_post2018)
terra::compareGeom(final_2011, ocean_mask_pre2018)

# Resolution and CRS checks
crs(final_2021)
crs(final_2021) == crs(final_2011)
res(final_2011)
res(final_2021)


```


